Variants of G\"{o}del's Ontological Proof \\ in a Natural Deduction Calculus

### 2016-11-07 Papers/InProgress/NaturalDeduction/GodProof-ND.tex
Fixed reviewer's comments.

{\ND} is a calculus for a (non-modal) higher-order logic without 
functional and boolean extensionality principles \citep{BenzmuellerBrownKohlhase}.
Adding the modal rules results in a calculus that is suitable for a basic rigid modal logic \textbf{K}, in the sense that {\NDK} is sound and complete relative to {\ND} extended with axiom K ($\nec(A\imp B)\imp (\nec A\imp \nec B)$) and the necessitation rule (which establishes that $\nec A$ is a theorem if $A$ is a theorem).

\begin{theorem}
\label{theorem:completeness}
{\NDK} is complete relative to {\ND} extended with axiom K and necessitation.
\end{theorem}
\begin{proof}
The necessitation rule can be immediately simulated with the $\nec_I$ rule. Axiom K can be derived in {\NDK} as shown below:

@@

\begin{theorem}
\label{theorem:soundness}
{\NDK} is sound relative to {\ND} extended with axiom K and necessitation.
\end{theorem}
\begin{proof}
It is necessary to show that {\NDK} proofs of the following form can be translated to proofs in {\ND} extended with the axiom K and the necessitation rule.

@@

\noindent
A translation to {\ND} extended with axiom K and necessitation is shown below for the case when $n=1$:

@@

The straightforward combinations of the quantifier rules of {\ND} either with the modal rules of {\NDK} or with axiom K and the necessitation rule are suitable for a higher-order modal logic where constants and variables are \emph{rigid}. From the point of view of a \emph{possible world semantics}, rigidity means that their interpretation is independent of the world in which they are being interpreted. Rigidity is tacitly assumed by most works investigating the ontological argument, and is explicitly assumed here. Nevertheless, it should be noted that its adequacy for the ontological argument has already been contested \citep{fitting}. Another assumption made here is that the quantification domains are constant (i.e. independent of the possible worlds). Neither G\"odel's manuscript nor Scott's manuscript reveal whether they use constant or varying domains, and this is also the case for many variants \citetext{e.g. \citealt{hajek}}. Nevertheless, some authors of variants of G\"odel's ontological argument do explicitly state a preference for varying domains. Our choice of constant domains is motivated by simplicity \citep[footnotes 11 and 14]{and90}. As is well-known, varying domain quantifiers (also known as \emph{actualistic} quantifiers) can be simulated by constant domain quantifiers (also known as \emph{possibilistic} quantifiers) guarded by a primitive \emph{existence} predicate\footnote{Formally, $\forall^e x. F[x] \equiv \forall x. E(x) \imp F[x]$ and $\exists^e x. F[x] \equiv \exists x. E(x) \wedge F[x]$.}; therefore, the formalization of ontological arguments with varying domain quantification would be feasible in the natural deduction calculus presented here as well.

@@

\noindent
\emph{Brouwer's reduction principle} is derivable in modal logic {\bf KB}, using axiom B. The proofs of G\"odel and Scott (Section \ref{sec:scott}) do not make direct use of this principle, but the new proof presented in Section \ref{sec:newproof} does.

@@

\begin{scriptsize}
\begin{prooftree}

\AXC{$ $} \RightLabel{1}
\UIC{$\pos \ex x. G(x)$} 
		\AXC{Lemma 1} \dashedLine
		\UIC{$\omega: \ex x. G(x) \imp \nec \ex x. G(x)$} \RightLabel{$\nec_I$}
		\UIC{$\nec (\ex x. G(x) \imp \nec \ex x. G(x)) $}
				\AXC{Distribution Principle}
			\BIC{$\pos \ex x. G(x) \imp \pos \nec \ex x. G(x)$} \RightLabel{$\imp_E$}
	\BIC{$\pos \nec \ex x. G(x)\scott$}
			\AXC{S5 Iteration Principle} \dashedLine
			\UIC{$\pos \nec \ex x. G(x) \imp \nec \ex x. G(x)$}\RightLabel{$\imp_E$}
		\BIC{$\nec \ex x. G(x)$} \RightLabel{$\imp_I^1$}
		\UIC{$\pos \ex z. G(z) \imp \nec \ex x. G(x)$}
\end{prooftree}
\end{scriptsize}

@@

In this section we present a new proof in {\bf KB}. This proof uses exactly the same axioms and definitions of Scott's proof shown in the previous sections, but it uses neither the axiom of reflexivity for equality nor {\bf S5}'s modal iteration principle. Instead, it relies only on Brouwer's reduction principle (for Theorem 3). The new proof is also shorter than Scott's proof.

@@

\begin{scriptsize}
\begin{prooftree}
\AXC{$ $} \RightLabel{3}
\UIC{$P(\rho)$} 
\AXC{$ $} \RightLabel{2}
\UIC{$\nec \neg \ex x.\rho(x) $} \RightLabel{$\nec_E$}
\UIC{$\omega: \neg \ex x.\rho(x) $} 
\AXC{$ $} \RightLabel{$1$}
\UIC{$\omega: \rho(x)$} \RightLabel{$\ex_I$}
\UIC{$\omega: \ex x.\rho(x) $} \RightLabel{$\neg_E$}
\BIC{$\omega:  \bot $}\RightLabel{$\imp_I^1$}
\UIC{$\omega:  \neg \rho(x) $} \RightLabel{$\imp_I$}
\UIC{$\omega:  \rho(x) \imp \neg \rho(x) $} \RightLabel{$\all_I$}
\UIC{$\omega:  \all x.(\rho(x) \imp \neg \rho(x)) $}  \RightLabel{$\nec_I$}
\UIC{$ \nec \all x.(\rho(x) \imp \neg \rho(x)) $}\RightLabel{$\wedge_I$}
\BIC{$ P(\rho) \wedge \nec \all x.(\rho(x) \imp \neg \rho(x))$}
\AXC{Axiom \ref{A2} for $\rho$ and $\lambda x. \neg \rho(x)$} \dashedLine
\UIC{$(P(\rho) \wedge \nec \all x.(\rho(x) \imp \neg \rho(x))) \imp P(\lambda x. \neg \rho(x))$} \RightLabel{$\imp_E$}
\BIC{$[\subproof_5]\qquad P(\lambda x. \neg \rho(x))$} \RightLabel{$\imp_E$}
\end{prooftree}
\end{scriptsize}

\begin{small}
\begin{prooftree}
\AXC{$\subproof_5$} \dashedLine
\UIC{$P(\lambda x. \neg \rho(x))$} %\RightLabel{$\imp_E$}
\AXC{Axiom 1} \dashedLine
\UIC{$\all \varphi. (P(\neg \varphi) \biimp \neg P(\varphi))$} \RightLabel{$\all_E$}
\UIC{$P(\lambda x. \neg \rho(x)) \biimp \neg P(\rho)$} \RightLabel{$\biimp_E$}
\BIC{$\neg P(\rho) $} 
\AXC{$ $} \RightLabel{3} 
\UIC{$P(\rho)$} \RightLabel{$\neg_E$}
\BIC{$\bot$} \RightLabel{$\imp_I^2$}
\UIC{$ \neg \nec \neg \ex x.\rho(x)$} \doubleLine %\RightLabel{definition of $\pos$}
\UIC{$ \pos \ex x.\rho(x)$}  \RightLabel{$\imp_I^3$}
\UIC{$ P(\rho) \imp \pos \ex x.\rho(x) $} \RightLabel{$\all_I$}
\UIC{$ \all \varphi. (P(\varphi) \imp \pos \ex x.\varphi(x)) $}

@@

\begin{small}
\begin{prooftree}
\AXC{$ $} \RightLabel{2}
\UIC{$\neg P(\rho)$} 
\AXC{Axiom 1} \dashedLine
\UIC{$\all \varphi. (P(\neg \varphi) \biimp \neg P(\varphi))$} \RightLabel{$\all_E$}
\UIC{$P(\lambda x. \neg \rho(x)) \biimp \neg P(\rho)$} \RightLabel{$\biimp_E$}
\BIC{$P(\lambda x. \neg \rho(x))$}
\AXC{$ $} \RightLabel{1}
\UIC{$\ex z. G(z)$}\RightLabel{$\ex_E$}
\UIC{$G(\gamma)$}  \dottedLine \RightLabel{D\ref{N:D1}}
\UIC{$\forall \varphi.(P(\varphi) \imp \varphi(\gamma))$}\RightLabel{$\forall_E$}
\UIC{$P(\lambda x. \neg \rho(x)) \imp \neg \rho(\gamma)$}\RightLabel{$\imp_E$}
\BIC{$\neg \rho(\gamma)$}
\AXC{$ $} \RightLabel{3}
\UIC{$\rho(\gamma)$} \RightLabel{$\neg_E$}
\BIC{$\bot$}\RightLabel{$\imp_I^2$}
\UIC{$\neg\neg P(\rho)$}\RightLabel{$\neg\neg_E$}
\UIC{$[\subproof_6] \qquad P(\rho)$}
\end{prooftree}
\end{small}

@@

%\AXC{$ $} \RightLabel{6}
%\UIC{$\rho(\gamma)$} \RightLabel{}
\AXC{$ \subproof_6$} \dashedLine
\UIC{$P(\rho)$} %\RightLabel{$\nec_E$}
\AXC{Axiom \ref{N:A4}} \dashedLine
\UIC{$\all \varphi.(P(\varphi) \imp \nec\; P(\varphi))$}\RightLabel{$\forall_E$}
\UIC{$P(\rho) \imp \nec\; P(\rho)$}\RightLabel{$\imp_E$}
\BIC{$\nec P(\rho)$}\RightLabel{$\imp_E$}%\RightLabel{$\imp_I^3$}
%\UIC{$[\subproof_7] \qquad \rho(\gamma) \imp \nec\; P(\rho)$}
%\UIC{$\rho(\gamma)\imp\Box P(\rho)$}\RightLabel{$\imp_E$}
%\BIC{$\nec P(\rho)$} \RightLabel{$\imp_E$}

\BIC{$\nec \forall y .(G(y)\imp \rho(y))$}\RightLabel{$\imp_I^6$}
\UIC{$\rho(\gamma) \imp \nec \forall y.(G(y) \imp \rho(y))$}\RightLabel{$\forall_I$}
\UIC{$\forall \psi.(\psi(\gamma) \imp \nec \forall y.(G(y) \imp \psi(y)))$}
\AXC{$ $} \RightLabel{1}
\UIC{$\ex z. G(z)$}\RightLabel{$\ex_E$}
\UIC{$G(\gamma)$}  \RightLabel{$\wedge_I$}
\BIC{$G(\gamma) \wedge \forall \psi.(\psi(\gamma) \imp \nec \forall y.(G(y) \imp \psi(y)))$}\dottedLine \RightLabel{D\ref{N:D2}}
\UIC{$[\subproof_7] \qquad \ess{G}{\gamma}$}
\end{prooftree}
\end{small}

\begin{small}
\begin{prooftree}
\AXC{$\subproof_7$} \dashedLine
\UIC{$\ess{G}{\gamma}$}
\AXC{Axiom \ref{A5}} \dashedLine
\UIC{$P(E)$}\dottedLine 
\AXC{$ $} \RightLabel{1}
\UIC{$\ex z. G(z)$}\RightLabel{$\ex_E$}
\UIC{$G(\gamma)$}\dottedLine \RightLabel{D\ref{N:D1}}
\UIC{$\forall \varphi.(P(\varphi) \imp \varphi(\gamma))$}\RightLabel{$\forall_E$}
\UIC{$P(E) \imp E(\gamma)$}\RightLabel{$\imp_E$}
\BIC{$E(\gamma)$}\dottedLine \RightLabel{D\ref{N:D3}}
\UIC{$ \all \varphi.(\ess{\varphi}{\gamma} \imp \nec \ex x.\varphi(x)) $}\RightLabel{$\all_E$}
\UIC{$ \ess{G}{\gamma} \imp \nec \ex x. G(x) $}\RightLabel{$\imp_E$}
\BIC{$\nec \ex x. G(x)$} \RightLabel{$\imp_I^1$}
\UIC{$\ex z. G(z) \imp \nec \ex x. G(x)$}
\end{prooftree}
\end{small}
\end{proof}

@@

\begin{small}
\begin{prooftree}
\AXC{ Distribution Principle} \dashedLine
\UIC{$\nec (\ex x. G(x)\imp \nec \ex x. G(x)) \imp (\pos \ex x. G(x)\imp \pos\nec \ex x. G(x))$}
\AXC{ Lemma \ref{N:L1} } \dashedLine
\UIC{$\ex x. G(x)\imp \nec \ex x. G(x)$}\RightLabel{$\nec_I$}
\UIC{$\nec(\ex x. G(x)\imp \nec \ex x. G(x))$}\RightLabel{$\imp_E$}
\BIC{$\pos \ex x. G(x)\imp \pos\nec \ex x. G(x)$}
\end{prooftree}
\end{small}
\begin{small}
\begin{prooftree}
\AXC{$\pos \ex x. G(x)\imp \pos\nec \ex x. G(x)$}
\AXC{ Corollary \ref{C1} } \dashedLine
\UIC{$\pos \ex x. G(x)$}\RightLabel{$\imp_E$}
\BIC{$\pos\nec \ex x. G(x)$}
\AXC{ Brouwer's Reduction Principle } \dashedLine
\UIC{$\pos\nec \ex x. G(x) \imp \ex x. G(x)$}\RightLabel{$\imp_E$}
\BIC{$ \ex x. G(x)$} \RightLabel{$\nec_I$}
\UIC{$\nec \ex x. G(x)$} 
\end{prooftree}
\end{small}

@@

Curiously, in the new proof of Theorem 3, the actual existence of a God-like being ($\ex x. G(x)$) is proven \emph{before} its necessity ($\nec \ex x. G(x)$). This contrasts with G\"odel's proof, where the actual existence is never actually proven (although it could, nevertheless, be derived as a trivial corollary by applying the modal axiom T).

@@

\begin{small}
\begin{prooftree}
\AXC{Theorem 2}\dashedLine
\UIC{$\all y.(G(y) \imp \ess{G}{y})$} \RightLabel{D2 }\dottedLine
\UIC{$\all y.(G(y) \imp G(y) \wedge \all \psi. (\psi(y) \imp \nec \all x. (G(x) \imp \psi(x))))$}\doubleLine
\UIC{$\all y.(G(y) \imp \all \psi. (\psi(y) \imp \nec \all x. (G(x) \imp \psi(x))))$} \doubleLine
%\UIC{$\all y.(G(y) \imp (A(y) \imp \nec \all x. (G(x) \imp A(x))))$} \RightLabel{$A$ is constant}
\UIC{$\all y.(G(y) \imp (A \imp \nec \all x. (G(x) \imp A)))$} \doubleLine
\UIC{$[\subproof_8] \qquad \ex y.G(y) \imp (A \imp \nec \all x. (G(x) \imp A))$}
\end{prooftree}
\end{small}

\begin{small}
\begin{prooftree}
\AXC{$ \subproof_8 $}\dashedLine
\UIC{$\ex y.G(y) \imp (A \imp \nec \all x. (G(x) \imp A))$}
\AXC{Theorem 3}\dashedLine
\UIC{$\nec \ex y.G(y)$} \RightLabel{$\nec_E$}
\UIC{$\ex y.G(y)$} \RightLabel{$\imp_E$}
\BIC{$ A \imp \nec \all x. (G(x) \imp A)$} \doubleLine
\UIC{$ A \imp \nec (\ex x. G(x) \imp A)$}
\AXC{$ $} \RightLabel{1}
\UIC{$A$} \RightLabel{$\imp_E$}
\BIC{$ \nec (\ex x. G(x) \imp A)$} \RightLabel{$\nec_E$}
\UIC{$ \omega:  \ex x. G(x) \imp A$} %\RightLabel{$\imp_E$}
\AXC{Theorem 3}\dashedLine
\UIC{$\nec \ex y.G(y)$} \RightLabel{$\nec_E$}
\UIC{$\omega: \ex x.G(x)$} \RightLabel{$\imp_E$}
\BIC{$\omega: A$} \RightLabel{$\nec_I$}
\UIC{$ \nec A$} \RightLabel{$\imp_I^1$}
   \UIC{$A\imp \nec A$}
\end{prooftree}
\end{small}

@@

\begin{theorem}
\label{A:T1}
Positive properties are possibly exemplified:
$$
\all \varphi. (P(\varphi) \imp \pos \ex x.\varphi(x))
$$
\end{theorem}
\begin{proof}
This proof is omitted, because any of the proofs of Theorem 1 in Sections \ref{sec:scott} and \ref{sec:newproof} would suffice to prove this theorem with only a small modification: the equivalence of axiom \ref{A1} would not need to be reduced to an implication by conjunction elimination. 
\end{proof}

@@

Because of Anderson's emendations to Axiom 1 and Definition 1, the proof of Lemma 1 has to be modified as well. The proof shown below uses the modal axiom T to eliminate the extra box modality introduced by Anderson
in his emended Definition 1.

\begin{lemma}
\label{A:L1}
If there is a God-like being, then there is a God-like being necessarily:
$$
\ex z. G(z) \imp \nec \ex x. G(x)
$$
\end{lemma}
\begin{proof} \hfill
\begin{prooftree}
\AXC{$ $} \RightLabel{1}
\UIC{$\ex z. G(z)$}\RightLabel{$\ex_E$}
\UIC{$[\subproof_9] \quad G(\gamma)$}
\end{prooftree}

\begin{prooftree}
\AXC{$\subproof_9$} \dashedLine
\UIC{$G(\gamma)$} \dottedLine \RightLabel{D\ref{A:D1}}
\UIC{$\forall \varphi. (P(\lambda x.\varphi(x)) \biimp \nec \varphi(\gamma))$}\RightLabel{$\forall_E$}
\UIC{$P(\lambda x.\rho(x)) \biimp \nec \rho(\gamma)$}
\AXC{$ $} \RightLabel{4}
\UIC{$\nec\rho(\gamma)$} \RightLabel{$\biimp_E$}
\BIC{$P(\rho)$} \RightLabel{$\forall_E$}

>>>>

{\ND} is a calculus for a (non-modal) higher-order logic without 
functional and boolean extensionality principles \citep{BenzmuellerBrownKohlhase}.
Adding the modal rules results in a calculus that is suitable for a basic rigid modal logic \textbf{K}, in the sense that {\NDK} is sound and complete relative to {\ND} extended with Axiom K ($\nec(A\imp B)\imp (\nec A\imp \nec B)$) and the necessitation rule (which establishes that $\nec A$ is a theorem if $A$ is a theorem).

\begin{theorem}
\label{theorem:completeness}
{\NDK} is complete relative to {\ND} extended with Axiom K and necessitation.
\end{theorem}
\begin{proof}
The necessitation rule can be immediately simulated with the $\nec_I$ rule. Axiom K can be derived in {\NDK} as shown below:

@@

\begin{theorem}
\label{theorem:soundness}
{\NDK} is sound relative to {\ND} extended with Axiom K and necessitation.
\end{theorem}
\begin{proof}
It is necessary to show that {\NDK} proofs of the following form can be translated to proofs in {\ND} extended with the Axiom K and the necessitation rule.

@@

\noindent
A translation to {\ND} extended with Axiom K and necessitation is shown below for the case when $n=1$:

@@

The straightforward combinations of the quantifier rules of {\ND} either with the modal rules of {\NDK} or with Axiom K and the necessitation rule are suitable for a higher-order modal logic where constants and variables are \emph{rigid}. From the point of view of a \emph{possible world semantics}, rigidity means that their interpretation is independent of the world in which they are being interpreted. Rigidity is tacitly assumed by most works investigating the ontological argument, and is explicitly assumed here. Nevertheless, it should be noted that its adequacy for the ontological argument has already been contested \citep{fitting}. Another assumption made here is that the quantification domains are constant (i.e. independent of the possible worlds). Neither G\"odel's manuscript nor Scott's manuscript reveal whether they use constant or varying domains, and this is also the case for many variants \citetext{e.g. \citealt{hajek}}. Nevertheless, some authors of variants of G\"odel's ontological argument do explicitly state a preference for varying domains. Our choice of constant domains is motivated by simplicity \citep[footnotes 11 and 14]{and90}. As is well-known, varying domain quantifiers (also known as \emph{actualistic} quantifiers) can be simulated by constant domain quantifiers (also known as \emph{possibilistic} quantifiers) guarded by a primitive \emph{existence} predicate\footnote{Formally, $\forall^e x. F[x] \equiv \forall x. E(x) \imp F[x]$ and $\exists^e x. F[x] \equiv \exists x. E(x) \wedge F[x]$.}; therefore, the formalization of ontological arguments with varying domain quantification would be feasible in the natural deduction calculus presented here as well.

@@

\noindent
\emph{Brouwer's reduction principle} is derivable in modal logic {\bf KB}, using Axiom B. The proofs of G\"odel and Scott (Section \ref{sec:scott}) do not make direct use of this principle, but the new proof presented in Section \ref{sec:newproof} does.

@@

		\AXC{Lemma 1} \dashedLine
		\UIC{$\omega: \ex x. G(x) \imp \nec \ex x. G(x)$} \RightLabel{$\nec_I$}
		\UIC{$\nec (\ex x. G(x) \imp \nec \ex x. G(x)) $} 
				\AXC{Distribution Principle}\dashedLine
				\UIC{$ \nec (\ex x. G(x) \imp \nec \ex x. G(x))\imp (\pos \ex x. G(x) \imp \pos \nec \ex x. G(x))$} \RightLabel{$\imp_E$}
			\BIC{$[\subproof_5] \qquad \pos \ex x. G(x) \imp \pos \nec \ex x. G(x)$} \RightLabel{$\imp_E$}

\begin{small}
\begin{prooftree}
\AXC{$ $} \RightLabel{1}
\UIC{$\pos \ex x. G(x)$} 
 \AXC{$\subproof_5$}\dashedLine
 	\UIC{$\pos \ex x. G(x) \imp \pos \nec \ex x. G(x)$} \RightLabel{$\imp_E$}
 \BIC{$\pos \nec \ex x. G(x)\scott$} 
			\AXC{S5 Iteration Principle} \dashedLine
			\UIC{$\pos \nec \ex x. G(x) \imp \nec \ex x. G(x)$}\RightLabel{$\imp_E$}
		\BIC{$\nec \ex x. G(x)$} \RightLabel{$\imp_I^1$}
		\UIC{$\pos \ex z. G(z) \imp \nec \ex x. G(x)$}
\end{prooftree}
\end{small}

@@

In this section we present a new proof in {\bf KB}. This proof uses exactly the same axioms and definitions of Scott's proof shown in the previous sections, but it uses neither the Axiom of reflexivity for equality nor {\bf S5}'s Modal iteration principle. Instead, it relies only on Brouwer's reduction principle (for Theorem 3). The new proof is also shorter than Scott's proof.

@@

\begin{scriptsize}
\begin{prooftree}
\AXC{$ $} \RightLabel{3}
\UIC{$P(\rho)$} 
\AXC{$ $} \RightLabel{2}
\UIC{$\nec \neg \ex x.\rho(x) $} \RightLabel{$\nec_E$}
\UIC{$\omega: \neg \ex x.\rho(x) $} 
\AXC{$ $} \RightLabel{$1$}
\UIC{$\omega: \rho(x)$} \RightLabel{$\ex_I$}
\UIC{$\omega: \ex x.\rho(x) $} \RightLabel{$\neg_E$}
\BIC{$\omega:  \bot $}\RightLabel{$\imp_I^1$}
\UIC{$\omega:  \neg \rho(x) $} \RightLabel{$\imp_I$}
\UIC{$\omega:  \rho(x) \imp \neg \rho(x) $} \RightLabel{$\all_I$}
\UIC{$\omega:  \all x.(\rho(x) \imp \neg \rho(x)) $}  \RightLabel{$\nec_I$}
\UIC{$ \nec \all x.(\rho(x) \imp \neg \rho(x)) $}\RightLabel{$\wedge_I$}
\BIC{$ P(\rho) \wedge \nec \all x.(\rho(x) \imp \neg \rho(x))$}
\AXC{Axiom \ref{A2} for $\rho$ and $\lambda x. \neg \rho(x)$} \dashedLine
\UIC{$(P(\rho) \wedge \nec \all x.(\rho(x) \imp \neg \rho(x))) \imp P(\lambda x. \neg \rho(x))$} \RightLabel{$\imp_E$}
\BIC{$[\subproof_6]\qquad P(\lambda x. \neg \rho(x))$} \RightLabel{$\imp_E$}
\end{prooftree}
\end{scriptsize}

\begin{small}
\begin{prooftree}
\AXC{$\subproof_6$} \dashedLine
\UIC{$P(\lambda x. \neg \rho(x))$} %\RightLabel{$\imp_E$}
\AXC{Axiom 1} \dashedLine
\UIC{$\all \varphi. (P(\neg \varphi) \biimp \neg P(\varphi))$} \RightLabel{$\all_E$}
\UIC{$P(\lambda x. \neg \rho(x)) \biimp \neg P(\rho)$} \RightLabel{$\biimp_E$}
\BIC{$\neg P(\rho) $} 
\AXC{$ $} \RightLabel{3} 
\UIC{$P(\rho)$} \RightLabel{$\neg_E$}
\BIC{$\bot$} \RightLabel{$\imp_I^2$}
\UIC{$ \neg \nec \neg \ex x.\rho(x)$} \doubleLine %\RightLabel{definition of $\pos$}
\UIC{$ \pos \ex x.\rho(x)$}  \RightLabel{$\imp_I^3$}
\UIC{$ P(\rho) \imp \pos \ex x.\rho(x) $} \RightLabel{$\all_I$}
\UIC{$ \all \varphi. (P(\varphi) \imp \pos \ex x.\varphi(x)) $}

@@

\begin{small}
\begin{prooftree}
\AXC{$ $} \RightLabel{2}
\UIC{$\neg P(\rho)$} 
\AXC{Axiom 1} \dashedLine
\UIC{$\all \varphi. (P(\neg \varphi) \biimp \neg P(\varphi))$} \RightLabel{$\all_E$}
\UIC{$P(\lambda x. \neg \rho(x)) \biimp \neg P(\rho)$} \RightLabel{$\biimp_E$}
\BIC{$P(\lambda x. \neg \rho(x))$}
\AXC{$ $} \RightLabel{1}
\UIC{$\ex z. G(z)$}\RightLabel{$\ex_E$}
\UIC{$G(\gamma)$}  \dottedLine \RightLabel{D\ref{N:D1}}
\UIC{$\forall \varphi.(P(\varphi) \imp \varphi(\gamma))$}\RightLabel{$\forall_E$}
\UIC{$P(\lambda x. \neg \rho(x)) \imp \neg \rho(\gamma)$}\RightLabel{$\imp_E$}
\BIC{$\neg \rho(\gamma)$}
\AXC{$ $} \RightLabel{3}
\UIC{$\rho(\gamma)$} \RightLabel{$\neg_E$}
\BIC{$\bot$}\RightLabel{$\imp_I^2$}
\UIC{$\neg\neg P(\rho)$}\RightLabel{$\neg\neg_E$}
\UIC{$[\subproof_7] \qquad P(\rho)$}
\end{prooftree}
\end{small}

@@

%\AXC{$ $} \RightLabel{6}
%\UIC{$\rho(\gamma)$} \RightLabel{}
\AXC{$ \subproof_7$} \dashedLine
\UIC{$P(\rho)$} %\RightLabel{$\nec_E$}
\AXC{Axiom \ref{N:A4}} \dashedLine
\UIC{$\all \varphi.(P(\varphi) \imp \nec\; P(\varphi))$}\RightLabel{$\forall_E$}
\UIC{$P(\rho) \imp \nec\; P(\rho)$}\RightLabel{$\imp_E$}
\BIC{$\nec P(\rho)$}\RightLabel{$\imp_E$}%\RightLabel{$\imp_I^3$}
%\UIC{$[\subproof_7] \qquad \rho(\gamma) \imp \nec\; P(\rho)$}
%\UIC{$\rho(\gamma)\imp\Box P(\rho)$}\RightLabel{$\imp_E$}
%\BIC{$\nec P(\rho)$} \RightLabel{$\imp_E$}

\BIC{$\nec \forall y .(G(y)\imp \rho(y))$}\RightLabel{$\imp_I^6$}
\UIC{$\rho(\gamma) \imp \nec \forall y.(G(y) \imp \rho(y))$}\RightLabel{$\forall_I$}
\UIC{$\forall \psi.(\psi(\gamma) \imp \nec \forall y.(G(y) \imp \psi(y)))$}
\AXC{$ $} \RightLabel{1}
\UIC{$\ex z. G(z)$}\RightLabel{$\ex_E$}
\UIC{$G(\gamma)$}  \RightLabel{$\wedge_I$}
\BIC{$G(\gamma) \wedge \forall \psi.(\psi(\gamma) \imp \nec \forall y.(G(y) \imp \psi(y)))$}\dottedLine \RightLabel{D\ref{N:D2}}
\UIC{$[\subproof_8] \qquad \ess{G}{\gamma}$}
\end{prooftree}
\end{small}

\begin{small}
\begin{prooftree}
\AXC{$\subproof_8$} \dashedLine
\UIC{$\ess{G}{\gamma}$}
\AXC{Axiom \ref{A5}} \dashedLine
\UIC{$P(E)$}\dottedLine 
\AXC{$ $} \RightLabel{1}
\UIC{$\ex z. G(z)$}\RightLabel{$\ex_E$}
\UIC{$G(\gamma)$}\dottedLine \RightLabel{D\ref{N:D1}}
\UIC{$\forall \varphi.(P(\varphi) \imp \varphi(\gamma))$}\RightLabel{$\forall_E$}
\UIC{$P(E) \imp E(\gamma)$}\RightLabel{$\imp_E$}
\BIC{$E(\gamma)$}\dottedLine \RightLabel{D\ref{N:D3}}
\UIC{$ \all \varphi.(\ess{\varphi}{\gamma} \imp \nec \ex x.\varphi(x)) $}\RightLabel{$\all_E$}
\UIC{$ \ess{G}{\gamma} \imp \nec \ex x. G(x) $}\RightLabel{$\imp_E$}
\BIC{$\nec \ex x. G(x)$} \RightLabel{$\imp_I^1$}
\UIC{$\ex z. G(z) \imp \nec \ex x. G(x)$}
\end{prooftree}
\end{small}
\end{proof}

@@

\begin{small}
\begin{prooftree}
\AXC{ Distribution Principle} \dashedLine
\UIC{$\nec (\ex x. G(x)\imp \nec \ex x. G(x)) \imp (\pos \ex x. G(x)\imp \pos\nec \ex x. G(x))$}
\AXC{ Lemma \ref{N:L1} } \dashedLine
\UIC{$\ex x. G(x)\imp \nec \ex x. G(x)$}\RightLabel{$\nec_I$}
\UIC{$\nec(\ex x. G(x)\imp \nec \ex x. G(x))$}\RightLabel{$\imp_E$}
\BIC{$[\subproof_9] \qquad \pos \ex x. G(x)\imp \pos\nec \ex x. G(x)$}
\end{prooftree}
\end{small}
\begin{small}
\begin{prooftree}
\AXC{$\subproof_9$}\dashedLine
\UIC{$\pos \ex x. G(x)\imp \pos\nec \ex x. G(x)$}
\AXC{ Corollary \ref{C1} } \dashedLine
\UIC{$\pos \ex x. G(x)$}\RightLabel{$\imp_E$}
\BIC{$\pos\nec \ex x. G(x)$}
\AXC{ Brouwer's Reduction Principle } \dashedLine
\UIC{$\pos\nec \ex x. G(x) \imp \ex x. G(x)$}\RightLabel{$\imp_E$}
\BIC{$ \ex x. G(x)$} \RightLabel{$\nec_I$}
\UIC{$\nec \ex x. G(x)$} 
\end{prooftree}
\end{small}

@@

Curiously, in the new proof of Theorem 3, the actual existence of a God-like being ($\ex x. G(x)$) is proven \emph{before} its necessity ($\nec \ex x. G(x)$). This contrasts with G\"odel's proof, where the actual existence is never actually proven (although it could, nevertheless, be derived as a trivial corollary by applying the modal Axiom T).

@@

\begin{small}
\begin{prooftree}
\AXC{Theorem 2}\dashedLine
\UIC{$\all y.(G(y) \imp \ess{G}{y})$} \RightLabel{D2 }\dottedLine
\UIC{$\all y.(G(y) \imp G(y) \wedge \all \psi. (\psi(y) \imp \nec \all x. (G(x) \imp \psi(x))))$}\doubleLine
\UIC{$\all y.(G(y) \imp \all \psi. (\psi(y) \imp \nec \all x. (G(x) \imp \psi(x))))$} \doubleLine
%\UIC{$\all y.(G(y) \imp (A(y) \imp \nec \all x. (G(x) \imp A(x))))$} \RightLabel{$A$ is constant}
\UIC{$\all y.(G(y) \imp (A \imp \nec \all x. (G(x) \imp A)))$} \doubleLine
\UIC{$[\subproof_{10}] \qquad \ex y.G(y) \imp (A \imp \nec \all x. (G(x) \imp A))$}
\end{prooftree}
\end{small}

\begin{small}
\begin{prooftree}
\AXC{$ \subproof_{10} $}\dashedLine
\UIC{$\ex y.G(y) \imp (A \imp \nec \all x. (G(x) \imp A))$}
\AXC{Theorem 3}\dashedLine
\UIC{$\nec \ex y.G(y)$} \RightLabel{$\nec_E$}
\UIC{$\ex y.G(y)$} \RightLabel{$\imp_E$}
\BIC{$ A \imp \nec \all x. (G(x) \imp A)$} \doubleLine
\UIC{$ A \imp \nec (\ex x. G(x) \imp A)$}
\AXC{$ $} \RightLabel{1}
\UIC{$A$} \RightLabel{$\imp_E$}
\BIC{$ \nec (\ex x. G(x) \imp A)$} \RightLabel{$\nec_E$}
\UIC{$ \omega:  \ex x. G(x) \imp A$} %\RightLabel{$\imp_E$}
\AXC{Theorem 3}\dashedLine
\UIC{$\nec \ex y.G(y)$} \RightLabel{$\nec_E$}
\UIC{$\omega: \ex x.G(x)$} \RightLabel{$\imp_E$}
\BIC{$\omega: A$} \RightLabel{$\nec_I$}
\UIC{$ \nec A$} \RightLabel{$\imp_I^1$}
   \UIC{$A\imp \nec A$}
\end{prooftree}
\end{small}

@@

\begin{theorem}
\label{A:T1}
Positive properties are possibly exemplified:
$$
\all \varphi. (P(\varphi) \imp \pos \ex x.\varphi(x))
$$
\end{theorem}
\begin{proof}
This proof is omitted, because any of the proofs of Theorem 1 in Sections \ref{sec:scott} and \ref{sec:newproof} would suffice to prove this theorem with only a small modification: the equivalence of Axiom \ref{A1} would not need to be reduced to an implication by conjunction elimination. 
\end{proof}

@@

Because of Anderson's emendations to Axiom 1 and Definition 1, the proof of Lemma 1 has to be modified as well.

\begin{lemma}
\label{A:L1}
If there is a God-like being, then there is a God-like being necessarily:
$$
\ex z. G(z) \imp \nec \ex x. G(x)
$$
\end{lemma}
\begin{proof} \hfill
\begin{prooftree}
\AXC{$ $} \RightLabel{1}
\UIC{$\ex z. G(z)$}\RightLabel{$\ex_E$}
\UIC{$[\subproof_{11}] \quad G(\gamma)$}
\end{prooftree}

\begin{prooftree}
\AXC{$\subproof_{11}$} \dashedLine
\UIC{$G(\gamma)$} \dottedLine \RightLabel{D\ref{A:D1}}
\UIC{$\forall \varphi. (P(\lambda x.\varphi(x)) \biimp \nec \varphi(\gamma))$}\RightLabel{$\forall_E$}
\UIC{$P(\lambda x.\rho(x)) \biimp \nec \rho(\gamma)$}
\AXC{$ $} \RightLabel{4}
\UIC{$\nec\rho(\gamma)$} \RightLabel{$\biimp_E$}
\BIC{$P(\rho)$} \RightLabel{$\forall_E$}

\AXC{Axiom \ref{A:A4}} \dashedLine
\UIC{$P(\rho)\imp \nec P(\rho)$} \RightLabel{$\imp_E$}
\BIC{$[\subproof_{12}] \quad \nec P(\rho)$} %\RightLabel{$\imp_E$}
\end{prooftree}
